---
title: "More detailed analysis of Rubisco fitness data"
author: 'SYSTEM: `r version[13]`'
date: 'DATE: `r Sys.time()`'
output:
  html_document:
    toc: yes
    df_print: paged
  pdf_document:
    toc: yes
  html_notebook:
    theme: cosmo
    toc: yes
    number_sections: yes
params:
  cpus: 2
  input_dir: ./
  meta: ''
---

```{r load-required-packages, include = FALSE}
library(ggplot2)
library(tidyverse)
library(DESeq2)
library(DescTools)
library(colorblindr) #https://github.com/clauswilke/colorblindr
library(tidyr)
library(ggrepel)
library(dplyr)
library(ggnewscale)
```

```{r load-required-datasets, message=FALSE}
count_matrix <- read_tsv("../output_new/prepare/all_counts.tsv")
count_matrix$Gene <- NULL
fitness_data <- read_tsv("../output_new/fitness/result.tsv")
```

# Prepare plot for pool

The distribution of mutant variants in the pool before and after freezing the mutant pool are plotted. The count matrix has to be brought into the correct format for the DESeq2 analysis and a design matrix created.

```{r prepare-count-matrix-DESeq2}
count_matrix_pool <- as.data.frame(data.frame(sgRNA=count_matrix$sgRNA, pool1=count_matrix$pool_t0_R1, pool2=count_matrix$pool_t0_R2, pool3=count_matrix$pool_t0_R3))
count_matrix_pool_t1 <- as.data.frame(data.frame(sgRNA=count_matrix$sgRNA, pool1=count_matrix$pool_t1_R1, pool2=count_matrix$pool_t1_R2, pool3=count_matrix$pool_t1_R3))

row.names(count_matrix_pool) <- count_matrix_pool$sgRNA
row.names(count_matrix_pool_t1) <- count_matrix_pool_t1$sgRNA
count_matrix_pool$sgRNA <- NULL
count_matrix_pool_t1$sgRNA <- NULL

design_matrix <- data.frame(group=c("pool", "pool", "pool"))
row.names(design_matrix) = c("pool1", "pool2", "pool3")
```

DESeq2 is used to normalize among the three replicates.

```{r normalize-using-DESeq2, message=FALSE}
ddsPool <- DESeqDataSetFromMatrix(
  countData = count_matrix_pool,
  colData = design_matrix,
  design = ~ 1)
ddsPool <- estimateSizeFactors(ddsPool)
counts_norm_ddsPool <- as.data.frame(counts(ddsPool, normalized=TRUE))

ddsPool_t1 <- DESeqDataSetFromMatrix(
  countData = count_matrix_pool_t1,
  colData = design_matrix,
  design = ~ 1)
ddsPool_t1 <- estimateSizeFactors(ddsPool_t1)
counts_norm_ddsPool_t1 <- as.data.frame(counts(ddsPool_t1, normalized=TRUE))
```

Using ggplot2, a plot showing mean and stdev of the pool samples is prepared

```{r plot-pool, message=FALSE}
df_plot <- data.frame(mut=row.names(counts_norm_ddsPool), mean_count=apply(counts_norm_ddsPool, 1, mean), stdev=apply(counts_norm_ddsPool,1,sd))
df_plot$mut <- factor(df_plot$mut, levels=c("WT", "mut4", "mut2", "mut5", "mut1", "K", "B", "I", "G", "E", "A", "2B4", "2I2", "A41", "2A1", "4x"))
p <- ggplot(df_plot) + geom_bar(aes(x=mut, y=mean_count), stat="identity") + geom_errorbar(aes(x=mut, ymin=mean_count-stdev, ymax=mean_count+stdev), width=0.4) + theme_light() + xlab("Mutant variant") + ylab("Mean of normalized read counts")
p
ggsave("../output_images/pool_distribution.pdf", plot=p)

df_plot <- data.frame(mut=row.names(counts_norm_ddsPool_t1), mean_count=apply(counts_norm_ddsPool_t1, 1, mean), stdev=apply(counts_norm_ddsPool_t1,1,sd))
df_plot$mut <- factor(df_plot$mut, levels=c("WT", "mut4", "mut2", "mut5", "mut1", "K", "B", "I", "G", "E", "A", "2B4", "2I2", "A41", "2A1", "4x"))
p <- ggplot(df_plot) + geom_bar(aes(x=mut, y=mean_count), stat="identity") + geom_errorbar(aes(x=mut, ymin=mean_count-stdev, ymax=mean_count+stdev), width=0.4) + theme_light() + xlab("Mutant variant") + ylab("Mean of normalized read counts")
p
ggsave("../output_images/pool_distribution_afterFreeze.pdf", plot=p)
```

Calculate the Gini coefficient of the mean pool samples. "A Gini coefficient of 0 reflects perfect equality, [...] while a Gini coefficient of 1 [...] reflects maximal inequality" (en.wikipedia.org). 
Gini coefficients for all samples are pretty low. (<0.16)

```{r Gini-index}
Gini(df_plot$mean_count, unbiased=FALSE)
Gini(count_matrix_pool$pool1, unbiased=FALSE)
Gini(count_matrix_pool$pool2, unbiased=FALSE)
Gini(count_matrix_pool$pool3, unbiased=FALSE)
```

# Prepare different plots to visualize fitness values

```{r plotting-functions, include=FALSE}
linetype_vect <- c("solid", "dashed", "dotted", "dotdash", "longdash", "twodash","solid", "dashed", "dotted", "dotdash", "longdash", "twodash", "solid", "dashed", "dotted", "dotdash")

lineplot_Rubisco <- function(df_to_plot, file_name, title_plot){
  df_to_plot$sgRNA <- factor(df_to_plot$sgRNA, levels=c("WT", "mut4", "mut2", "mut5", "mut1", "K", "B", "I", "G", "E", "A", "2B4", "2I2", "A41", "2A1", "4x"))
  plot_p <- ggplot(df_to_plot, aes(x=time, y=log2FoldChange, color=sgRNA)) + geom_line(aes(linetype=sgRNA)) + theme_light() + ggtitle(title_plot) + xlab("Time (generations)") + ylab("log$_2$FC") + scale_linetype_manual(values=linetype_vect) + coord_cartesian(ylim=c(-2.75, +4.1), xlim=c(0,10))
  ggsave(file_name, plot=plot_p, width=18, height=18, units="cm")
  return(plot_p)
}


lineplot_gray_Rubisco <- function(df_to_plot, file_name, title_plot){
  df_to_plot$sgRNA <- factor(df_to_plot$sgRNA, levels=c("WT", "mut4", "mut2", "mut5", "mut1", "K", "B", "I", "G", "E", "A", "2B4", "2I2", "A41", "2A1", "4x"))
  plot_p <- ggplot(df_to_plot, aes(x=time, y=log2FoldChange, color=sgRNA, alpha=sgRNA)) + geom_line(linewidth=1.2) + theme_light() + ggtitle(title_plot) + xlab("Time (generations)") + ylab("Log2FC") + scale_linetype_manual(values=linetype_vect) + scale_color_manual(values=c("WT"="#000000ff", "K"="lightgray", "mut4"="#e59f02ff", "mut2"="#70b1e3ff", "mut5"="#449b76ff", "mut1"="#ba80a3ff", "A"="lightgray", "B"="lightgray", "E"="lightgray", "G"="lightgray", "I"="lightgray", "A41"="lightgray", "2A1"="lightgray", "2B4"="lightgray", "2I2"="lightgray", "4x"="#d75c00ff")) + scale_alpha_manual(values=c(c("WT"=1, "K"=0.7, "mut1"=1, "mut2"=1, "mut4"=1, "mut5"=1, "A"=0.7, "B"=0.7, "E"=0.7, "G"=0.7, "I"=0.7, "A41"=0.7, "2A1"=0.7, "2B4"=0.7, "2I2"=0.7, "4x"=1))) + coord_cartesian(ylim=c(-2.75, +4.1), xlim=c(0,10)) # + ylim(-2.75, +4.1)
  ggsave(file_name, plot=plot_p, width=18, height=18, units="cm")
  return(plot_p)
}

#p <- ggplot(combined_df, aes(x=time, y=mean, ymin=mean-margin, ymax=mean+margin, fill=strain, colour=strain, linetype=strain)) + geom_line() + geom_ribbon(alpha=0.25, linetype=0) + theme_light() + scale_y_log10() + scale_colour_manual(name="", values=colour_v) + scale_fill_manual(name="", values=colour_v) 
#p

# https://support.bioconductor.org/p/80725/#80729
lineplot_CVinterval_gray_Rubisco <- function(df_to_plot, file_name, title_plot){
  df_to_plot$sgRNA <- factor(df_to_plot$sgRNA, levels=c("WT", "mut4", "mut2", "mut5", "mut1", "K", "B", "I", "G", "E", "A", "2B4", "2I2", "A41", "2A1", "4x"))
  df_to_plot$conf <- qnorm(0.975)*df_to_plot$lfcSE
  plot_p <- ggplot(df_to_plot, aes(x=time, y=log2FoldChange, ymin=log2FoldChange-conf, ymax=log2FoldChange+conf)) + new_scale("alpha") + scale_color_manual(values=c("WT"="#000000ff", "K"="lightgray", "mut4"="#e59f02ff", "mut2"="#70b1e3ff", "mut5"="#449b76ff", "mut1"="#ba80a3ff", "A"="lightgray", "B"="lightgray", "E"="lightgray", "G"="lightgray", "I"="lightgray", "A41"="lightgray", "2A1"="lightgray", "2B4"="lightgray", "2I2"="lightgray", "4x"="#d75c00ff")) + scale_fill_manual(values=c("WT"="#000000ff", "K"="lightgray", "mut4"="#e59f02ff", "mut2"="#70b1e3ff", "mut5"="#449b76ff", "mut1"="#ba80a3ff", "A"="lightgray", "B"="lightgray", "E"="lightgray", "G"="lightgray", "I"="lightgray", "A41"="lightgray", "2A1"="lightgray", "2B4"="lightgray", "2I2"="lightgray", "4x"="#d75c00ff")) + scale_alpha_manual(values=c(c("WT"=1, "K"=0.7, "mut1"=1, "mut2"=1, "mut4"=1, "mut5"=1, "A"=0.7, "B"=0.7, "E"=0.7, "G"=0.7, "I"=0.7, "A41"=0.7, "2A1"=0.7, "2B4"=0.7, "2I2"=0.7, "4x"=1))) + geom_line(aes(alpha=sgRNA, color=sgRNA)) + new_scale("linetype") + new_scale("alpha") + scale_alpha_manual(values=c(c("WT"=0.1, "K"=0, "mut1"=0.1, "mut2"=0.1, "mut4"=0.1, "mut5"=0.1, "A"=0, "B"=0, "E"=0, "G"=0, "I"=0, "A41"=0, "2A1"=0, "2B4"=0, "2I2"=0, "4x"=0.1))) + geom_ribbon(linetype=0, aes(alpha=sgRNA, fill=sgRNA, color=sgRNA)) + theme_light() + ggtitle(title_plot) + xlab("Time (generations)") + ylab("Log2FC")  + coord_cartesian(ylim=c(-2.75, +4.1), xlim=c(0,10))# + ylim(-2.75, +4.1)
  ggsave(file_name, plot=plot_p, width=18, height=18, units="cm")
  return(plot_p)
}

barplot_Rubisco <- function(df_to_plot, file_name, title_plot){
  df_to_plot <- unique(df_to_plot[c("sgRNA", "fitness")])
  df_to_plot$sgRNA <- factor(df_to_plot$sgRNA, levels=c("WT", "mut4", "mut2", "mut5", "mut1", "K", "B", "I", "G", "E", "A", "2B4", "2I2", "A41", "2A1", "4x"))
  plot_bar <- ggplot(df_to_plot, aes(y=fitness, x=sgRNA)) + geom_bar(stat="identity")+ theme_light() + ggtitle(title_plot) + xlab("Mutant variant") + ylab("Fitness")
  ggsave(file_name, plot=plot_bar)
  return(plot_bar)
}
```

Plot for constant light, 5% CO2, N2 conditions:

```{r HL_5N2}
df_HL_5N2 <- subset(fitness_data, fitness_data$condition=="HL_5N2")
a <- lineplot_Rubisco(df_to_plot=df_HL_5N2, file_name="../output_images/lineplot/HL_5N2_lineplot.pdf", title_plot="Constant light, 5% CO2, N2")
a
a <- lineplot_gray_Rubisco(df_to_plot=df_HL_5N2, file_name="../output_images/lineplot_gray/HL_5N2_lineplot.pdf", title_plot="Constant light, 5% CO2, N2")
a
a <- barplot_Rubisco(df_HL_5N2, "../output_images/barplots/HL_5N2_barplot.pdf", "Constant light, 5% CO2, N2")
a
a <- lineplot_CVinterval_gray_Rubisco(df_to_plot=df_HL_5N2, file_name="../output_images/lineplot_gray/HL_5N2_lineplot_ribbon.pdf", title_plot="Constant light, 5% CO2, N2")
a
```


```{r HL_5O2}
HL_5O2 <- subset(fitness_data, fitness_data$condition=="HL_5O2")
a <- lineplot_Rubisco(HL_5O2, "../output_images/lineplot/HL_5O2_lineplot.pdf", "Constant light, 5% CO2, ambient O2")
a
a <- lineplot_gray_Rubisco(HL_5O2, "../output_images/lineplot_gray/HL_5O2_lineplot.pdf", "Constant light, 5% CO2, ambient O2")
a
a <- barplot_Rubisco(HL_5O2, "../output_images/barplots/HL_5O2_barplot.pdf", "Constant light, 5% CO2, ambient O2")
a
a <- lineplot_CVinterval_gray_Rubisco(df_to_plot=HL_5O2, file_name="../output_images/lineplot_gray/HL_5O2_lineplot_ribbon.pdf", title_plot="Constant light, 5% CO2, N2")
a
```


```{r HL_pool}
pool <- subset(fitness_data, fitness_data$condition=="pool")
a <- lineplot_Rubisco(pool, "../output_images/lineplot/pool_lineplot.pdf", "pool")
a
a <- barplot_Rubisco(pool, "../output_images/barplots/pool_barplot.pdf", "pool")
a
```

```{r LD_N2}
LD_N2 <- subset(fitness_data, fitness_data$condition=="LD_N2")
a <- lineplot_Rubisco(LD_N2, "../output_images/lineplot/light-dark_lineplot.pdf", "Light-dark cycles, 5% CO2, N2")
a
a <- lineplot_gray_Rubisco(LD_N2, "../output_images/lineplot_gray/light-dark_lineplot.pdf", "Light-dark cycles, 5% CO2, N2")
a
a <- lineplot_CVinterval_gray_Rubisco(df_to_plot=LD_N2, file_name="../output_images/lineplot_gray/light-dark_lineplot_ribbon.pdf", title_plot="Light-dark cycles, 5% CO2, N2")
a
a <- barplot_Rubisco(LD_N2, "../output_images/barplots/light-dark_barplot.pdf", "Light-dark cycles, 5% CO2, N2")
a
```


```{r barplot of all}
fitness_data_to_plot <- unique(fitness_data[c("sgRNA", "fitness", "condition")])
fitness_data_to_plot <- subset(fitness_data_to_plot, fitness_data_to_plot$condition=="HL_5N2" | fitness_data_to_plot$condition=="HL_5O2" | fitness_data_to_plot$condition=="LD_N2")
fitness_data_to_plot$sgRNA <- factor(fitness_data_to_plot$sgRNA, levels=c("WT", "mut4", "mut2", "mut5", "mut1", "K", "B", "I", "G", "E", "A", "2B4", "2I2", "A41", "2A1", "4x"))
p <- ggplot(fitness_data_to_plot, aes(y=fitness, x=sgRNA, fill=condition)) + geom_bar(stat="identity", position="dodge")+ theme_light() + xlab("Mutant variant") + ylab("Fitness") + scale_fill_manual(values=c("HL_5N2"="#999999ff", "HL_5O2"="#d45d00ff", "LD_N2"="#0072b1ff"))# scale_fill_OkabeIto()
p
ggsave("../output_images/barplot_differentconditions.pdf", plot=p, height=11, width=18, units="cm")
cvd_grid(p)
```
```{r barplot-of-all-scaled}
fitness_data_to_plot <- unique(fitness_data[c("sgRNA", "fitness", "condition")])
fitness_data_to_plot <- subset(fitness_data_to_plot, fitness_data_to_plot$condition=="HL_5N2" | fitness_data_to_plot$condition=="HL_5O2" | fitness_data_to_plot$condition=="LD_N2")
fitness_data_to_plot[fitness_data_to_plot$condition=="HL_5N2",]$fitness <- scale(fitness_data_to_plot[fitness_data_to_plot$condition=="HL_5N2",]$fitness)[,1]
fitness_data_to_plot[fitness_data_to_plot$condition=="HL_5O2",]$fitness <- scale(fitness_data_to_plot[fitness_data_to_plot$condition=="HL_5O2",]$fitness)[,1]
fitness_data_to_plot[fitness_data_to_plot$condition=="LD_N2",]$fitness <- scale(fitness_data_to_plot[fitness_data_to_plot$condition=="LD_N2",]$fitness)[,1]
fitness_data_to_plot$sgRNA <- factor(fitness_data_to_plot$sgRNA, levels=c("WT", "mut4", "mut2", "mut5", "mut1", "K", "B", "I", "G", "E", "A", "2B4", "2I2", "A41", "2A1", "4x"))
p <- ggplot(fitness_data_to_plot, aes(y=fitness, x=sgRNA, fill=condition)) + geom_bar(stat="identity", position="dodge")+ theme_light() + xlab("Mutant variant") + ylab("Z-scaled fitness values") + scale_fill_manual(values=c("HL_5N2"="#999999ff", "HL_5O2"="#d45d00ff", "LD_N2"="#0072b1ff"))# scale_fill_OkabeIto()
p
ggsave("../output_images/barplot_differentconditions_scaled.pdf", plot=p, height=11, width=18, units="cm")
```

# Scatter plots

```{r scatter-plots}
red_fitness <- unique(fitness_data[c("sgRNA", "condition", "fitness")])
wide_fitness <- pivot_wider(red_fitness, names_from=condition, values_from=c(fitness))
```

```{r scatter_N_O}
lm_N2_O2 <- lm(HL_5O2~HL_5N2, wide_fitness)
correlation <- cor.test(wide_fitness$HL_5O2, wide_fitness$HL_5N2, method = 'pearson')
correlation
correlation <- cor.test(wide_fitness$HL_5O2, wide_fitness$HL_5N2, method = 'spearman')
correlation
scat <- ggplot(wide_fitness, aes(x=HL_5N2, y=HL_5O2, label=sgRNA)) + geom_point(color="darkgray") + theme_light() + geom_text_repel() + geom_abline(intercept=lm_N2_O2$coefficients[1],slope=lm_N2_O2$coefficients[2],linetype="dashed",color="black") + xlim(-2.75, 3.5) + ylim(-2.75, 3.5) + xlab("Fitness constant light, 5% CO2, N2") + ylab("Fitness constant light, 5% CO2, ambient air")
scat
ggsave("../output_images/scatterplots/scatter_N2_O2.pdf", scat, width=18, height=18, units="cm")
```

```{r scatter_N_LD}
scat <- ggplot(wide_fitness, aes(x=HL_5N2, y=LD_N2, label=sgRNA)) + geom_point(color="darkgray") + theme_light()  + geom_text_repel() + xlab("Fitness constant light, 5% CO2, N2") + ylab("Fitness light-dark cycles, 5% CO2, N2")
scat
ggsave("../output_images/scatterplots/scatter_N2_LD.pdf", scat, width=18, height=18, units="cm")
```

```{r scatter_EVmut}
EVmut <- read.table("../input/2024-03-26_predictions_combinations_smallSubset.tsv", sep="\t", header=TRUE)
combo_sgRNA <- read.table("../input/combo_mutName.csv", header=TRUE)
EVmut_sgRNA <- left_join(EVmut, combo_sgRNA)
wide_fitness_EVmut <- left_join(wide_fitness, EVmut_sgRNA)
scat <- ggplot(wide_fitness_EVmut, aes(x=HL_5N2, y=Score, label=sgRNA)) + geom_point(color="darkgray") + theme_light()  + geom_text_repel() + xlab("Fitness constant light, 5% CO2, N2") + ylab("EVmutation score")
scat
ggsave("../output_images/scatterplots/HL_5N2_EVmut.pdf", scat, width=18, height=18, units="cm")
```

```{r scatter_LD_EVmut}
scat <- ggplot(wide_fitness_EVmut, aes(x=LD_N2, y=Score, label=sgRNA)) + geom_point(color="darkgray") + theme_light()  + geom_text_repel() + xlab("Fitness light-dark cycles, 5% CO2, N2") + ylab("EVmutation score")
scat
ggsave("../output_images/scatterplots/LDN2_EVmut.pdf", scat)
```

```{r scatter_numMut}
wide_fitness_EVmut$numberMut <- c(3,3,3,4,2,3,2,2,2,2,2,1,1,1,1,0)
scat <- ggplot(wide_fitness_EVmut, aes(x=LD_N2, y=numberMut, label=sgRNA)) + geom_point(color="darkgray") + theme_light()  + geom_text_repel() + xlab("Fitness light-dark cycles, 5% CO2, N2") + ylab("Number mutations")
scat
ggsave("../output_images/scatterplots/LDN2_numMut.pdf", scat)
scat <- ggplot(wide_fitness_EVmut, aes(x=HL_5N2, y=numberMut, label=sgRNA)) + geom_point(color="darkgray") + theme_light()  + geom_text_repel() + xlab("Fitness constant light, 5% CO2, N2") + ylab("Number mutations")
scat
ggsave("../output_images/scatterplots/N2_numMut.pdf", scat)
```


```{r linear-regressions}
summary(lm(HL_5N2 ~ Score, wide_fitness_EVmut))
summary(lm(HL_5N2 ~ numberMut, wide_fitness_EVmut))
summary(lm(LD_N2 ~ Score, wide_fitness_EVmut))
```


# Session Info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

